Supplementary material

### 2025-07-11 03-methods.qmd
Update carpeta resubmission y varios (archivos no renderizados)

pacman::p_load(tidyverse, 
               sjmisc, 
               sjPlot, 
               lme4, 
               here, 
               performance,
               influence.ME,
               marginaleffects,
               MLMusingR,
               texreg, 
               ggdist,
               misty,
               kableExtra,
               ggalluvial, 
               shadowtext,
               MetBrewer,
               patchwork,
               sjlabelled,
               summarytools)

@@

This study draws on data from the Chilean Longitudinal Social Survey (ELSOC), a panel study conducted annually from 2016 to 2023. The survey assesses how individuals think, feel, and behave regarding social conflict and cohesion in Chile. ELSOC employs a probabilistic, stratified, clustered, multistage sampling design covering major urban centers (Santiago, Valparaíso, and Concepción) as well as smaller cities. The target population includes men and women aged 18 to 75 who are habitual residents of private dwellings.

The survey has been conducted every year since 2016, except in 2020, when it was suspended due to the COVID-19 pandemic. The first wave included 2,927 participants from both northern and southern regions, covering 77% of the national population and 93% of the urban population, with a response rate of 62.4% [@elsoc_estudio_2022]. This study uses six waves: 2016, 2017, 2018, 2019, 2022, and 2023. The 2021 wave was excluded because a reduced version of the questionnaire omitted key variables of interest. Between waves 1 and 6, panel attrition reached 40%, resulting in a final two-level sample comprising N = 8,643 observations nested within N = 1,687 individuals. Longitudinal weights are applied to adjust for both the sampling design and potential biases from systematic non-response. Further details on sampling, attrition, and weighting procedures are available at https://coes.cl/encuesta-panel/, and the dataset is publicly accessible at https://dataverse.harvard.edu/dataverse/elsoc.

@@

**Market justice preferences**: The dependent variable in this study is preferences for market justice. This construct is operationalized through three items that capture how strongly individuals justify conditioning access to social services---healthcare, pensions, and education--- basen on individual income. Specifically, the justification of inequality in healthcare is assessed by the question: "Is it fair in Chile that people with higher incomes can access better healthcare than people with lower incomes?" The same question is posed for pensions and education. In all cases, respondents indicate their level of agreement on a five-point Likert scale ranging from 1 ("strongly disagree") to 5 ("strongly agree"). Additionally, we include a composite measure of "market justice preferences", calculated as the average of these three items ($\alpha$ = 0.84). This index ranges from 1 to 5, with higher values indicating stronger preferences for market justice (see @tbl-summary1).

@@

**Perception of economic inequality**: The main independent variable refers to the perception of economic inequality, measured through the perceived wage gap [@castillo_cual_2009; @hadler_why_2005; @gijsberts_thelegitimation_1999]. This measure is derived from the salary gap between the perceived salaries of jobs at opposite ends of the occupational hierarchy. Specifically, it relies on the division between the perceived salary of a large-company president and that of an unskilled worker [@Castillo2011]. Higher values of this term indicate a greater perception of economic inequality between occupations located at the extremes of the status continuum. This measure includes a logarithmic term in order to adjust income magnitudes (usually fewer cases with high income):

@@

Sociodemographic and attitudinal variables are included to control for potential composition effects in the population. In terms of sociodemographic characteristics, we incorporate per capita household income quintile, educational level (1=Less than Universitary, 2=Universitary), age (in years), and sex (1=Male, 2=Female), which have been previously shown to influence market justice preferences significantly [@lindh_public_2015; @castillo_socialization_2024]. Regarding attitudinal variables, we include political identification (1=Left, 2=Center, 3=Right, 4=No identification) and subjective social status (ranging from 1 to 10) as they may affect the relationship between market justice preferences, perceptions of inequality, and meritocracy [@schneider_poverty_2015].

>>>>

pacman::p_load(tidyverse, 
               sjmisc, 
               sjPlot, 
               lme4, 
               here, 
               performance,
               marginaleffects,
               texreg, 
               ggdist,
               kableExtra,
               ggalluvial, 
               shadowtext,
               MetBrewer,
               patchwork,
               sjlabelled,
               summarytools)

@@

This study draws on data from the Chilean Longitudinal Social Survey (ELSOC), a nationally representative panel study of the urban adult population in Chile, conducted annually between 2016 and 2023. Designed to examine individuals’ attitudes, emotions, and behaviors regarding social conflict and cohesion, ELSOC employs a probabilistic, stratified, clustered, and multistage sampling design covering both major urban centers and smaller cities. The sampling frame was proportionally stratified into six categories of urban population size (e.g., large and small cities), followed by the random selection of households within 1,067 city blocks. The target population includes men and women aged 18 to 75 who are habitual residents of private dwellings.

The survey has been conducted every year since 2016, except in 2020, when it was suspended due to the COVID-19 pandemic. This study uses six waves: 2016, 2017, 2018, 2019, 2022, and 2023. The 2021 wave was excluded because a reduced version of the questionnaire omitted key variables of interest. Between waves 1 and 6, panel attrition reached 40%, resulting in a final two-level sample comprising N = 8,643 observations nested within N = 1,687 individuals. Longitudinal weights are applied to adjust for both the sampling design and potential biases from systematic non-response. Further details on sampling, attrition, and weighting procedures are available at https://coes.cl/encuesta-panel/, and the dataset is publicly accessible at https://dataverse.harvard.edu/dataverse/elsoc.

@@

**Market justice preferences**: the dependent variable in this study is preferences for market justice. This construct is operationalized through three items that capture how strongly individuals justify conditioning access to core services---healthcare, pensions, and education--- basen on individual income. Specifically, the justification of inequality in healthcare is assessed by the question: "Is it fair in Chile that people with higher incomes can access better healthcare than people with lower incomes?" The same question is posed for pensions and education. In all cases, respondents indicate their level of agreement on a five-point Likert scale ranging from 1 ("strongly disagree") to 5 ("strongly agree"). Additionally, we include a composite measure of "market justice preferences", calculated as the average of these three items ($\alpha$ = 0.84). This index ranges from 1 to 5, with higher values indicating stronger preferences for market justice (see @tbl-summary1).

@@

**Perception of economic inequality**: this variable is measured through the perceived wage gap [@castillo_cual_2009; @hadler_why_2005; @gijsberts_thelegitimation_1999]. This measure is derived from the salary gap between the perceived salaries of jobs at opposite ends of the occupational hierarchy. Specifically, it relies on the division between the perceived salary of a large-company president and that of an unskilled worker [@Castillo2011]. Higher values of this term indicate a greater perception of economic inequality between occupations located at the extremes of the status continuum. This measure includes a logarithmic term in order to adjust income magnitudes (usually fewer cases with high income):

@@

Sociodemographic and attitudinal variables are included to control for potential composition effects in the population. In terms of sociodemographic characteristics, we incorporate per capita household income quintile, educational level (1=Less than Universitary, 2=Universitary), age (in ranges), and sex (1=Male, 2=Female), which have been previously shown to influence market justice preferences significantly [@lindh_public_2015; @castillo_socialization_2024]. Regarding attitudinal variables, we include political identification (1=Left, 2=Center, 3=Right, 4=No identification) and subjective social status (measured through a scale from 1 to 10) as they may affect the relationship between market justice preferences, perceptions of inequality, and meritocracy [@schneider_poverty_2015]. Descriptive statistics for the control variables can be found in the supplementary material.

### 2025-07-11 04-analisis.qmd
Update carpeta resubmission y varios (archivos no renderizados)

pacman::p_load(tidyverse, 
               sjmisc, 
               sjPlot, 
               lme4, 
               here, 
               performance,
               influence.ME, 
               marginaleffects,
               MLMusingR,
               texreg, 
               ggdist,
               misty,
               kableExtra,
               ggalluvial, 
               shadowtext,
               MetBrewer,
               patchwork,
               sjlabelled,
               summarytools,
               interactions)

@@

## Descriptive

@@

Regarding the main dependent and independent variables of this study, @fig-meanchange depicts their average changes over the years. We observe an increase in the average level of market justice preferences in the most recent waves, which begins in 2019. The highest average consistently appears for perceived economic inequality, although this variable shows a downward trend of roughly one point over time. Interestingly, while perceptions of economic inequality declined in the latest measurements (2022–2023), market justice preferences increased. The meritocracy measures remain stable, though the perception that individuals are rewarded for intelligence is slightly higher than the perception that they are rewarded for their effort.

```{r}
#| label: fig-meanchange
#| out-width: '100%'
#| fig-asp: 0.8
#| fig-cap: "Change in the mean of market justice preferences, economic inequality perception, and meritocracy (2016-2023)"
#| fig-cap-location: top
#| echo: false
#| results: asis

@@

df_pond %>% 
  select(ola, mjp, perc_inequality, merit_effort, merit_talent) %>% 
  mutate_at(.vars = 2:5, .funs = ~ as.numeric(.)) %>% 
  group_by(ola) %>% 
  summarise_all(~survey_mean(., vartype = "ci")) %>% 
  pivot_longer(cols = -ola,
               names_to = "temp",
               values_to = "valor") %>%
  mutate(
    ci = case_when(
      str_ends(temp, "_low") ~ "ic_low",
      str_ends(temp, "_upp") ~ "ic_upp",
      TRUE                   ~ "mean"
    ),
    variable = str_remove(temp, "_low|_upp")
  ) %>%
  select(ola, variable, ci, valor) %>%
  pivot_wider(
    names_from  = ci,
    values_from = valor
  ) %>% 
  mutate(variable = case_when(variable == "mjp" ~ "Market justice preferences",
                              variable == "perc_inequality"  ~ "Inequality perception",
                              variable == "merit_effort"     ~ "Merit: Effort",
                              variable == "merit_talent"     ~ "Merit: Talent"),
         variable = factor(variable, levels = c("Market justice preferences",
                                      "Inequality perception",
                                      "Merit: Effort",
                                      "Merit: Talent")
                              )) %>% 
  ggplot(aes(x = ola, y = mean, group = variable)) +
  geom_point(aes(shape=variable, color=variable), size = 3.5) +
  geom_line(aes(color = variable), linewidth = 0.8) +
  geom_errorbar(aes(ymin = ic_low, ymax = ic_upp, color = variable),
                width = 0.1) +
  scale_color_manual(values = c("#023858", # azul oscuro
                                "#0571B0", # azul intermedio
                                "#F4A582", # salmón
                                "#B2182B"  # rojo profundo
                                         )) +
  scale_shape_manual(values=c(15, 16, 17, 18)) +
  labs(y = "Mean",
       x = "Wave",
       color = NULL,
       shape = NULL,
       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (N obs = 8,643; N groups = 1,687)") +
  theme_ggdist() +
  theme(legend.position = "top",
        text = element_text(size = 12))

@fig-matrix presents the correlation matrix for the main variables in the latest wave (2023). Overall, the coefficients range from low to moderate. The association between market justice preferences and economic inequality perception is negative but small and statistically significant ($r$ = -0.05, $p$ < .05). By contrast, market justice preferences positively and significantly correlate with the two meritocracy variables ($r$ = 0.15, $p$ < .01; $r$ = 0.13, $p$ < .01). Perceived economic inequality has negative and nonsignificant correlations with both meritocracy perceptions ($r$ = -0.04, $p$ > .05; $r$ = -0.04, $p$ > .05). Finally, the two meritocracy variables exhibit a strong positive association with each other ($r$ = 0.69, $p$ < .01).

```{r}
#| label: fig-matrix
#| out-width: '90%'
#| fig-asp: 1
#| fig-cap: "Correlation matrix of the main variables for the last wave (2023)"
#| fig-cap-location: top
#| echo: false
#| results: asis

M <- df_study1 %>%
  filter(ola == 6) %>% 
  select(mjp, perc_inequality, merit_effort, merit_talent) %>% 
  mutate_all(.funs = ~ as.numeric(.)) %>% 
  cor(use = "pairwise.complete.obs", method = "pearson")

diag(M) <- NA

rownames(M) <- c("A. Marjet justice preferences",
                 "B. Inequality gap perception",
                 "C. People are rewarded for their efforts ",
                 "D. People are rewarded for their intelligence")

#set Column names of the matrix
colnames(M) <-c("(A)", "(B)","(C)","(D)")

#Plot the matrix using corrplot
corrplot::corrplot(M,
                   method = "color",
                   addCoef.col = "black",
                   type = "upper",
                   tl.col = "black",
                   col = colorRampPalette(c("#E16462", "white", "#0D0887"))(12),
                   bg = "white",
                   na.label = "-")

@@

Models 3 and 4 incorporate the within-group effects (WE) of the primary independent variables, capturing how individual changes in these variables over time shape the dependent variable. The results in Model 3 suggest that the within effect of perceived economic inequality is negative and statistically significant ($p$ < .001). Specifically, each one-point increase in an individual’s perception of economic inequality between waves is associated with a 0.034 point decrease in market justice preferences. Model 4 shows that meritocratic beliefs operate in the opposite direction. An upward shift in the perception that effort is rewarded exerts a positive within effect ($\beta$ = 0.071, $p$ < .001), and a parallel increase in the perception that intelligence and ability are rewarded is likewise associated with higher market-justice preferences ($\beta$ = 0.028, $p$ < .05). Taken together, these results suggest that individuals who increasingly perceive meritocracy—whether through effort or talent—tend to hold stronger market justice preferences.

@@

@@

@tbl-interactions1 examines whether perceived economic inequality moderates the effect of meritocratic beliefs on market justice preferences. Contrary to our hypothesis, the interaction terms in the within-person specification (Models 8 and 9) do not reach statistical significance. However, in the between-person specification (Models 10 and 11), the interaction terms are statistically significant, suggesting that perceptions of economic inequality meaningfully alter the effect of meritocracy on support for market-based allocation of social services. These positive effects indicate that, as perceptions of inequality increase, the positive association between meritocratic perceptions and market justice preferences becomes stronger. The marginal effects of this interaction are illustrated in @fig-interact, which shows that the effect of meritocratic perceptions on market justice preferences intensifies as perceived economic inequality rises.

>>>>

pacman::p_load(tidyverse, 
               sjmisc, 
               sjPlot, 
               lme4, 
               here, 
               performance,
               marginaleffects,
               texreg, 
               ggdist,
               kableExtra,
               ggalluvial, 
               shadowtext,
               MetBrewer,
               patchwork,
               sjlabelled,
               summarytools,
               interactions,
               cowplot)

@@

## Descriptive statistics

@@

Regarding the main dependent and independent variables of this study, @fig-meanchange depicts their standardized average changes over the years. We observe an increase in the average level of market justice preferences in the most recent waves, which begins in 2019. The highest average appears for perceived economic inequality in 2016, although this variable shows a downward trend over time, with the exception of 2019, which can be associated with the “social crisis” in the country. Interestingly, while perceptions of economic inequality declined in the latest measurements (2022–2023), market justice preferences increased. The meritocracy measures remain stable, though the perception that individuals are rewarded for effort is slightly higher than the perception that they are rewarded for their intelligence

```{r}
#| label: fig-meanchange
#| out-width: '100%'
#| fig-asp: 0.8
#| fig-cap: "Change in the standarized mean of market justice preferences, economic inequality perception, and meritocracy (2016-2023)"
#| fig-cap-location: top
#| echo: false
#| results: asis

@@

df_pond <- df_pond %>% 
  select(ola, mjp, perc_inequality, merit_effort, merit_talent) %>% 
  mutate_at(.vars = 2:5, .funs = ~ as.numeric(.)) %>% 
  mutate(
    mjp_z            = as.numeric(scale(mjp)),
    perc_inequality_z = as.numeric(scale(perc_inequality)),
    merit_effort_z    = as.numeric(scale(merit_effort)),
    merit_talent_z    = as.numeric(scale(merit_talent))
  )

df_pond %>% 
  select(ola, mjp_z, perc_inequality_z, merit_effort_z, merit_talent_z) %>% 
  group_by(ola) %>% 
  summarise_all(~survey_mean(., vartype = "ci")) %>% 
  pivot_longer(cols = -ola,
               names_to = "temp",
               values_to = "valor") %>%
  mutate(
    ci = case_when(
      str_ends(temp, "_low") ~ "ic_low",
      str_ends(temp, "_upp") ~ "ic_upp",
      TRUE                   ~ "mean"
    ),
    variable = str_remove(temp, "_low|_upp")
  ) %>%
  select(ola, variable, ci, valor) %>%
  pivot_wider(
    names_from  = ci,
    values_from = valor
  ) %>% 
  mutate(variable = case_when(variable == "mjp_z" ~ "Market justice preferences",
                              variable == "perc_inequality_z"  ~ "Inequality perception",
                              variable == "merit_effort_z"     ~ "Merit: Effort",
                              variable == "merit_talent_z"     ~ "Merit: Talent"),
         variable = factor(variable, levels = c("Market justice preferences",
                                      "Inequality perception",
                                      "Merit: Effort",
                                      "Merit: Talent")
                              )) %>% 
  ggplot(aes(x = ola, y = mean, group = variable)) +
  geom_point(aes(shape=variable, color=variable), size = 3.5) +
  geom_line(aes(color = variable), linewidth = 0.8) +
  geom_errorbar(aes(ymin = ic_low, ymax = ic_upp, color = variable),
                width = 0.1) +
  scale_color_manual(values = c("#023858", # azul oscuro
                                "#0571B0", # azul intermedio
                                "#F4A582", # salmón
                                "#B2182B"  # rojo profundo
                                         )) +
  scale_shape_manual(values=c(15, 16, 17, 18)) +
  labs(y = "Mean (Z-score)",
       x = "Wave",
       color = NULL,
       shape = NULL,
       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (N obs = 8,643; N groups = 1,687)") +
  theme_ggdist() +
  theme(legend.position = "top",
        text = element_text(size = 12))

@@

Models 3 and 4 incorporate the within-group effects (WE) of the primary independent variables, capturing how individual changes in these variables over time shape the dependent variable. The results in Model 3 suggest that the within effect of perceived economic inequality is negative and statistically significant ($p$ < .001). Specifically, each one-point increase in an individual’s perception of economic inequality between waves is associated with a 0.034 point decrease in market justice preferences. Model 4 shows that meritocratic perceptions operate in the opposite direction. An upward shift in the perception that effort is rewarded exerts a positive within effect ($\beta$ = 0.071, $p$ < .001), and a parallel increase in the perception that intelligence and ability are rewarded is likewise associated with higher market-justice preferences ($\beta$ = 0.028, $p$ < .05). Taken together, these results suggest that individuals who increasingly perceive meritocracy—whether through effort or talent—tend to hold stronger market justice preferences.

@@

ggeffects::ggpredict(model = m9.1,
                     terms = c("merit_effort_mean",
                               "perc_inequality_mean [min, max]"),
                     ci_level = .95
                     ) %>% 
  plot()

predictions(m9.1,
            condition = c("merit_effort_mean", "perc_inequality_mean"),
            newdata = datagrid(merit_effort_mean = df_study1$merit_effort_mean,
                               perc_inequality_mean = seq(min(df_study1$perc_inequality_mean), max(df_study1$perc_inequality_mean), by= 0.1)),
            re.form = NA) %>% 
  ggplot(aes(y =estimate,x=merit_effort_mean, fill=perc_inequality_mean, color=perc_inequality_mean)) +
  geom_ribbon(alpha=0.8,size=1,linetype="solid",color="black") +
  geom_line(size=1,color="black") +
  geom_point(size=2,color="black")

@@

@tbl-interactions1 examines whether perceived economic inequality moderates the effect of meritocratic perceptions on market justice preferences. Contrary to our hypothesis, the interaction terms in the within-person specification (Models 8 and 9) do not reach statistical significance. However, in the between-person specification (Models 10 and 11), the interaction terms are statistically significant, suggesting that perceptions of economic inequality meaningfully alter the effect of meritocracy on support for market-based allocation of social services. These positive effects indicate that, as perceptions of inequality increase, the positive association between meritocratic perceptions and market justice preferences becomes stronger. The marginal effects of this interaction are illustrated in @fig-interact, which shows that the effect of meritocratic perceptions on market justice preferences intensifies as perceived economic inequality rises.

### 2025-07-11 05-discussion.qmd
Update carpeta resubmission y varios (archivos no renderizados)

This study tests several key hypotheses concerning the relationships between perceptions of economic inequality, meritocratic beliefs, and attitudes toward market justice in Chile. It posits that higher perceived inequality and stronger meritocratic beliefs both independently contribute to greater support for market-based distributions of social goods. Furthermore, the study examines how perceived economic inequality moderates the influence of meritocratic beliefs on market justice preferences. Additionally, it considers how major social and political mobilizations, such as those occurring around 2019 and subsequent years, may have weakened these relationships, reflecting increased critical perspectives on systemic inequalities and market mechanisms. The subsequent discussion interprets the empirical findings in light of these hypotheses, considering their implications for understanding attitudes toward inequality and social justice in a context marked by profound social change.

@@

Regarding the second hypothesis, the results reinforced that meritocratic perceptions ---particularly those emphasizing individual effort and talent--- were associated with stronger support for market-based distribution systems. Individuals who viewed success as primarily earned through effort were more inclined to justify existing inequalities, seeing them as outcomes of individual merit rather than systemic injustice. This aligned with research indicating that meritocratic narratives served as moral justifications that reinforced societal stratification [@Castillo2012a_justice; @hoyt_mindsets_2023]. These beliefs functioned symbolically to legitimize structural disparities, reducing support for redistributive measures by framing inequality as fair and deserved. The findings suggested that these perceptions bolstered the social hierarchy by fostering acceptance of inequality as a reflection of individual virtue rather than structural failure, which was especially pertinent in neoliberal contexts like Chile's, where market logic heavily influenced social attitudes [@canalesceron_sujeto_2021].

The third hypothesis concerned the moderating role of perceptions of inequality on the relationship between meritocracy and market justice attitudes. The analysis indicated that the positive association between meritocratic beliefs and support for market-based distribution tended to weaken when perceptions of economic inequality were particularly high. Although this moderation is not significant for within-subject effects (i.e. over time), when taking into account between-subjects differences we found a positive effect. This appears as contrary to our initial hypothesis, as it would mean that effect of meritocratic perceptions becomes stronger as perception of inequality increases. Nevertheless, we need to be cautious about this interpretation, as the main effect of inequality perception on market justice is negative. Therefore, as perceived inequality become less negative (i.e. more positive), or closer to zero, then the meritocratic perception has an stronger positive relationship on market justice preferences. This could mean that meritocratic perceptions play an stronger role in justification of market justice inequalities for those who are less capable or less prone to perceive inequality.

Finally, the fourth hypothesis addressed the impact of recent social movements, particularly from 2019 onwards, on the relationship between meritocratic beliefs, perceptions of inequality, and attitudes toward market justice. The results suggested that these movements contributed to a more critical stance toward the fairness of existing distributional mechanisms. Post-mobilization, the positive relationship between meritocratic beliefs and support for inequality appeared attenuated, indicating that collective action and public discourse during this period fostered questioning of the legitimacy of systemic inequalities justified solely through meritocratic narratives. These social mobilizations challenged dominant narratives, emphasizing structural barriers and systemic injustices, which in turn weakened the association between individual effort and deservingness. This is aligned with literature emphasizing that social protests act as catalysts for transforming societal attitudes by disrupting normative beliefs and encouraging more critical evaluations of institutional legitimacy [@tilly_contentious_2015]. The findings underscored the importance of socio-political context in shaping and reshaping attitudes toward inequality and justice, highlighting how collective action could temporarily or permanently alter the normative foundations that sustained support for market justice.

>>>>

This study tests several key hypotheses concerning the relationships between perceptions of economic inequality, meritocratic perceptions, and attitudes toward market justice in Chile. It posits that higher perceived inequality and stronger meritocratic perceptions both independently contribute to greater support for market-based distributions of social goods. Furthermore, the study examines how perceived economic inequality moderates the influence of meritocratic perceptions on market justice preferences. Additionally, it considers how major social and political mobilizations, such as those occurring around 2019 and subsequent years, may have weakened these relationships, reflecting increased critical perspectives on systemic inequalities and market mechanisms. The subsequent discussion interprets the empirical findings in light of these hypotheses, considering their implications for understanding attitudes toward inequality and social justice in a context marked by profound social change.

@@

Regarding the second hypothesis, the results reinforced that meritocratic perceptions ---particularly those emphasizing individual effort and talent--- were associated with stronger support for market-based distribution systems. Individuals who viewed success as primarily earned through effort were more inclined to justify existing inequalities, seeing them as outcomes of individual merit rather than systemic injustice. This aligned with research indicating that meritocratic narratives served as moral justifications that reinforced societal stratification [@Castillo2012a_justice; @hoyt_mindsets_2023]. These perceptions functioned symbolically to legitimize structural disparities, reducing support for redistributive measures by framing inequality as fair and deserved. The findings suggested that these perceptions bolstered the social hierarchy by fostering acceptance of inequality as a reflection of individual virtue rather than structural failure, which was especially pertinent in neoliberal contexts like Chile's, where market logic heavily influenced social attitudes [@canalesceron_sujeto_2021].

The third hypothesis concerned the moderating role of perceptions of inequality on the relationship between meritocracy and market justice attitudes. The analysis indicated that the positive association between meritocratic perceptions and support for market-based distribution tended to weaken when perceptions of economic inequality were particularly high. Although this moderation is not significant for within-subject effects (i.e. over time), when taking into account between-subjects differences we found a positive effect. This appears as contrary to our initial hypothesis, as it would mean that effect of meritocratic perceptions becomes stronger as perception of inequality increases. Nevertheless, we need to be cautious about this interpretation, as the main effect of inequality perception on market justice is negative. Therefore, as perceived inequality become less negative (i.e. more positive), or closer to zero, then the meritocratic perception has an stronger positive relationship on market justice preferences. This could mean that meritocratic perceptions play an stronger role in justification of market justice inequalities for those who are less capable or less prone to perceive inequality.

Finally, the fourth hypothesis addressed the impact of recent social movements, particularly from 2019 onwards, on the relationship between meritocratic perceptions, perceptions of inequality, and attitudes toward market justice. The results suggested that these movements contributed to a more critical stance toward the fairness of existing distributional mechanisms. Post-mobilization, the positive relationship between meritocratic perceptions and support for inequality appeared attenuated, indicating that collective action and public discourse during this period fostered questioning of the legitimacy of systemic inequalities justified solely through meritocratic narratives. These social mobilizations challenged dominant narratives, emphasizing structural barriers and systemic injustices, which in turn weakened the association between individual effort and deservingness. This is aligned with literature emphasizing that social protests act as catalysts for transforming societal attitudes by disrupting normative beliefs and encouraging more critical evaluations of institutional legitimacy [@tilly_contentious_2015]. The findings underscored the importance of socio-political context in shaping and reshaping attitudes toward inequality and justice, highlighting how collective action could temporarily or permanently alter the normative foundations that sustained support for market justice.

### 2025-07-11 06-conclusion.qmd
Update carpeta resubmission y varios (archivos no renderizados)

This study examined the complex interplay between perceptions of economic inequality, meritocratic beliefs, and attitudes toward market justice in Chile from 2016 to 2023, drawing on longitudinal data from the ELSOC survey. By exploring how subjective assessments and social contexts influence support for redistribution and market-based resource allocation, the research offers different elements that contribute to the understanding of the normative foundations underpinning social justice attitudes in a highly unequal, neoliberal environment.

The findings confirm that higher perceptions of economic inequality are associated with less support for market justice attitudes, this is, the belief that it is fine that those with higher income have better social services such as education, pensions and health. Meritocratic beliefs strongly reinforce support for market justice, serving as moral justifications for structural inequalities. Additionally, perceptions of inequality moderate the influence of meritocratic beliefs, such that as the awareness of disparities are attenuated (i.e. less negative) the role of meritocratic beliefs on existing market-based allocations become stronger. Notably, the social mobilizations of 2019 and subsequent periods seem to have contributed to shifting attitudes, fostering greater skepticism of systemic fairness and reducing the legitimizing power of meritocratic narratives.

This research advances the extant literature by integrating subjective perceptions with social and political contexts to explain attitudes toward economic inequality and distributional justice. While previous studies primarily focused on objective measures or individual characteristics, this work emphasizes the dynamic and interactional nature of perceptions and beliefs over time. It highlights the importance of socio-political upheavals in reshaping normative attitudes, underscoring the role of collective action in challenging entrenched narratives of meritocracy and fairness. The longitudinal approach provides a deeper temporal perspective on how societal events could influence individual beliefs and preferences.

>>>>

This study examined the complex interplay between perceptions of economic inequality, meritocratic perceptions, and attitudes toward market justice in Chile from 2016 to 2023, drawing on longitudinal data from the ELSOC survey. By exploring how subjective assessments and social contexts influence support for redistribution and market-based resource allocation, the research offers different elements that contribute to the understanding of the normative foundations underpinning social justice attitudes in a highly unequal, neoliberal environment.

The findings confirm that higher perceptions of economic inequality are associated with less support for market justice attitudes, this is, the belief that it is fine that those with higher income have better social services such as education, pensions and health. Meritocratic perceptions strongly reinforce support for market justice, serving as moral justifications for structural inequalities. Additionally, perceptions of inequality moderate the influence of meritocratic perceptions, such that as the awareness of disparities are attenuated (i.e. less negative) the role of meritocratic perceptions on existing market-based allocations become stronger. Notably, the social mobilizations of 2019 and subsequent periods seem to have contributed to shifting attitudes, fostering greater skepticism of systemic fairness and reducing the legitimizing power of meritocratic narratives.

This research advances the extant literature by integrating subjective perceptions with social and political contexts to explain attitudes toward economic inequality and distributional justice. While previous studies primarily focused on objective measures or individual characteristics, this work emphasizes the dynamic and interactional nature of perceptions and beliefs over time. It highlights the importance of socio-political upheavals in reshaping normative attitudes, underscoring the role of collective action in challenging entrenched narratives of meritocracy and fairness. The longitudinal approach provides a deeper temporal perspective on how societal events could influence individual perceptions and preferences.

### 2025-07-11 07-supplementary-material.qmd
Update carpeta resubmission y varios (archivos no renderizados)

---
title: "Supplementary material"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  docx:
    number-sections: true
  pdf: 
    geometry: margin=0.5cm
    template-partials: 
      - title.tex
    keep-tex: true
    number-sections: true
editor: source
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
--- 
```{r}
#| echo: false
#| message: false
#| warning: false

@@

pacman::p_load(tidyverse, 
               sjmisc, 
               sjPlot, 
               lme4, 
               here, 
               performance,
               influence.ME, 
               marginaleffects,
               MLMusingR,
               texreg, 
               ggdist,
               misty,
               kableExtra,
               ggalluvial, 
               shadowtext,
               MetBrewer,
               patchwork,
               sjlabelled,
               summarytools,
               interactions)

@@

>>>>

---
title: "Supplementary material"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  docx:
    number-sections: true
  pdf: 
    geometry: margin=0.5cm
    template-partials: 
      - title.tex
    keep-tex: true
    number-sections: true
editor: source
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
editor_options: 
  chunk_output_type: console
--- 
```{r}
#| echo: false
#| message: false
#| warning: false

@@

pacman::p_load(tidyverse, 
               sjmisc, 
               sjPlot, 
               lme4, 
               here, 
               performance,
               texreg, 
               ggdist,
               kableExtra,
               shadowtext,
               MetBrewer,
               patchwork,
               sjlabelled,
               summarytools)

@@

```{r}
#| label: tbl-summary-controls
#| tbl-cap: "Descriptive statistics for control variables for the first wave (2016)"
#| tbl-cap-location: top
#| results: asis
#| echo: false

t2 <- df_study1 %>% 
  filter(ola == "2016") %>% 
  select(educ,
         quintil1,
         sex,
         age,
         ess, 
         ideo)

t2$educ <- sjlabelled::set_label(t2$educ, "Educational level")
t2$quintil1 <- sjlabelled::set_label(t2$quintil1, "Household income quintile (per capita)")
t2$sex <- sjlabelled::set_label(t2$sex, "Sex")
t2$age <- sjlabelled::set_label(t2$age, "Age")
t2$ess <- sjlabelled::set_label(t2$ess, "Subjective social status")
t2$ideo <- sjlabelled::set_label(t2$ideo, "Political identification")

df<-dfSummary(t2,
               plain.ascii = FALSE,
               style = "multiline",
               tmp.img.dir = "/tmp",
               graph.magnif = 0.75,
               headings = F,  # encabezado
               varnumbers = F, # num variable
               labels.col = T, # etiquetas
               na.col = F,    # missing
               graph.col = F, # plot
               valid.col = T, # n valido
               col.widths = c(30,10,10,10))

df$Variable <- NULL # delete variable column

print(df)
```
